Formalizing the Edmonds-Karp Algorithm

### 2016-03-03 paper/root.tex
Rephrasing + improvement up to and excluding benchmark section

\section{Introduction}
The Ford-Fulkerson algorithm~\cite{FF56} is one of the important results of the graph theory and is used to find a solution to the maximum flow problem in flow networks. Many important mathematical problems like the maximum-bipartite-matching problem, the edge-disjoint-paths problem, the circulation-demand problem, and many other scheduling and resource allocating problems can be reduced to the maximum flow problem. Hence, the Ford-Fulkerson algorithm has major application in the field of mathematical optimization.

Despite its importance, no formalization of the Ford-Fulkerson algorithm has been developed in modern proof assistants. The only similar work that the authors are aware of is formalization of the Ford-Fulkerson algorithm in Mizar~\cite{Lee05}. This formalization defines and proves correctness of the algorithm at the level of graph manipulations without providing concrete implementation of the algorithm. Providing such an implementation is specially important, as it could provide the basis for many verfied programs with practical importance.

(TODO: Here I have mentioned the Edmonds-Karp for the first time.) This paper present a formalization of the correctness proof of the Ford-Fulkerson method in Isabelle/HOL~\cite{NPW02}. Our formalization is based on the informal proof presented in the book \emph{Introduction to algorithms}~\cite{CLRS09}. We then provide a verified implementation of the Edmons-Karp's algorithm~\cite{EK72}. This algorithm instantiates the Ford-Fulkerson method with breadth-first-search and provides a polynomial algorithm for solving the maximum flow problem in generic case--- maximum flow on networks with real edge capacities.

Isabelle/HOL provides some automation for generating the code corresponding to a verified algorithm, however, in order to use the code generating features, one needs to do the formalization with executability in mind. Being limited only to executable concepts in the formalization has the disadvantage of cluttering the proofs with implementation details. Such approach makes the proofs more complicated, and may even render proofs of medium complex algorithms unmanageable.

(TODO: I am not sure about the level of details that I describe refinement in introduction, co-relate this with the refinement section) One solution for the aforementioned problem is refinement~\cite{Wirth71}. In order to generate executable code of an algorithm using refinement, we first formulate the algorithm on an abstract level. The abstract version of the algorithm has a clean correctness proof as it only captures the idea behind the algorithm. Next, we refine the abstract definition of the algorithm towards an executable implementation in possibly multiple refinement steps. During each step, we only need to prove the correctness of the implementation of a particular abstract concept. Hence, we can be sure about the correctness of the resulting executable program as each refinement step preserves the correctness.

(TODO: check the citation order and description) There are several approaches to data refinement in Isabelle/HOL. We will be using the Isabelle Refinement Framework~\cite{LaTu12,La12}, which has been used for proving more complex results such as formalized implementation of Hopcroft’s DFA-minimization algorithm~\cite{LaTu12}. Given an algorithm phrased over abstract concepts like sets and maps, it automatically synthesizes a concrete, executable algorithm and the corresponding refinement theorem. To make it applicable for the development of actual algorithms, Autoref is integrated with the Isabelle Refinement Framework~\cite{La12} and the Isabelle Collection Framework~\cite{LL10}.

\section{A short background on the Ford-Fulkerson method}
% \begin{comment}
%   GOAL: 
%     1) Remind the reader of FoFu. The educated reader should have the feeling:
%       Yes, now I recall FoFu, and understand (again) what it does.
%       
%     2) Set the field for further, more detailed descriptions in next section.
%   
% 
%   Short description of Network: Finite graph over nodes V, 
%     edge (u,v) annotated with capacity c(u,v).
%     Assuming distinct nodes s and t. 
%     Note on additional assumptions: 
%         No parallel edges, s only outgoing, t only incoming, all nodes on path from s to t. 
%         Can transform any Network to match these assumptions, preserving the flow \cite{???}.
%         
%   s-t Flow: Annotation of edges with values, such that: 
%     1) Values smaller than capacities.
%     2) Kirchhoff law: sum of incoming flows = sum of outgoing flows, for all nodes except s t
%   
%   Value of flow: Incoming flow - outgoing flow of s
%   
%   Max-Flow problem. 
%     Rpt. why important?
%   
%   Cuts, intuition of min-cut >= max-flow
%   
%   Theorem: Min-Cut = Max-Flow
%   Proven via augmenting flow in residual graph. 
%     Present 1,2,3 of min-cut max-flow equivalences
%   
%   This yields an algorithm to compute max-flows, if we can compute flows in residual graph.
%     --> Simple way: Augmenting flow via augmenting path. ==> FoFu-scheme. 
%       Termination? In General: Only for integer capacities.
%       Edmonds-Karp: Shortest augmenting path. Always terminates! Complexity: O(|E||V|) outer loop iterations, BFS requires O|E|.
%   \end{comment}
A flow network is a weighted digraph where the set of vertices (nodes) $V$ is finite, and each edge $(u, v)$ has a real capacity $c (u, v) \geq 0$. There are two distinct vertices $s$ (source) and $t$ (sink) in the flow network. In this case the network is called an $s$-$t$ network. To make the formalization simple, some additional assumptions are added to the definition as following: 1) The source only has outgoing edges while the sink only has incoming edges; 2) If the flow network contains an edge $(u, v)$ then there is no edge $(v, u)$ in the reverse direction; and 3) every vertex of the flow network must be on a path from $s$ to $t$. Although these assumptions seems restrictive, any network may be transformed such that it satisfies all the aforementioned properties~\cite{CLRS09}.

An $s$-$t$ flow on a weighted digraph is a function $\colon V \times V \rightarrow \mathbb{R}$ which satisfies the following conditions. 1) \emph{Capacity constraint}: the flow of each edge is a non-negative value which is smaller or equal to the edge capacity; 2) \emph{Conservation constraint}: For all vertices except $s$ and $t$, the sum of flows over incoming edges of the vertex is equal to the sum of flows over outgoing edges of the vertex.

The value of an $s$-$t$ flow f is denoted by $|f|$, and is computed by subtracting the sum of the flows over all incoming edges of $s$ from the sum of the flows over all outgoing edges of $s$. Given an $s$-$t$ network $G$, the maximum flow problem involves finding a flow $f$ in $G$ such that the value of $f$ is greater or equal to the value of any other $s$-$t$ flow in $G$~\footnote{Note that the value of the flow in our model of a flow network equals sum of flows over edges going out of $s$. Because the model prohibits edges coming into the source.}.

One of the important results in mathematical optimization is the correlation between flows and cuts in flow networks. An $s$-$t$ cut in a flow network with source $s$ and sink $t$ is a partition of vertices that puts $s$ and $t$ in different subsets. Then the capacity of a cut is defined as sum of the edge capacities over all edges going from the source's side to the sink's side. Consider a valid flow in a flow network $G$. For any cut in $G$, the flow that crosses between the two subsets of the cut cannot exceed the capacity of the cut. So cuts define an upper bound for any flow in the flow network. The Ford-Fulkerson theorem tightens this bound and states that the value of the maximum flow is equal to the capacity of the minimum cut.

The Ford-Fulkerson algorithm is the direct consequence of the Ford-Fulkerson theorem. It solves the maximum flow problem in a greedy approach. Assume an instance of the maximum flow problem in an $s$-$t$ network. The algorithm starts with a zero flow: $f(e) = 0$ for all the edges $e$. In each iteration of the algorithm the value of $f$ is increased by pushing flow along a path from $s$ to $t$ up to the limits imposed by the available edge capacities. The process would continue until no more paths can be found to push additional flow from $s$ to $t$.

The total flow value is increased by manipulating the flow values along different edges. However, the flow along specific edges might decrease during this process. In order to perform these operations, the Ford-Fulkerson algorithm defines the residual graph of the flow network. Intuitively, the residual graph corresponding to a flow in a given flow network, has the same vertices as the flow network and consists of edges with capacities representing how flow can be changed along the edges of the flow network.

%Assume a flow $f$ in an $s$-$t$ network with capacity function $c$. An edge $(u, v)$ of the network can admit an additional amount of flow equal to the edge’s capacity minus the flow along that edge. If this value is positive, then a $(u, v)$ edge is added to the residual graph with a residual capacity $c_f (u, v) = c (u, v) - f (u, v)$. So, the residual graph only contains those edges of the original flow network that have left-over capacities. The residual graph may also contain some edges that do not exist in the original flow network. For each edge $(u, v)$ in the flow network which has a positive flow value, a $(v, u)$ edge is added to the residual graph with the residual capacity $c_f (v, u) = f (u, v)$. The edge $(v, u)$ indicates the possibility for pushing the flow backward along the edge $(u, v)$ in the original flow network.

(TODO: there is a shorter informal description available as comment. Adapt that one if required) There are two kinds of edges in the residual graph: 1) those of the residual graph that present the edges of the flow network with left-over capacities; and 2) those of the residual graph that do not occur in the flow network, and indicate the possibility for pushing the flow backward. Formally speaking, consider a flow $f$ on an $s$-$t$ network $G = (V, E)$ with the capacity function $c$. A residual capacity function $c_f \colon V \times V \rightarrow \mathbb{R}$ is defined as following:
\[ c_f (u, v) = 
	\begin{cases}
	c (u, v) - f(u, v) \hfill & \text{ if $(u, v) \in E$} \\
	f (v, u) \hfill & \text{ if $(v, u) \in E$} \\	
	0 \hfill & \text{ otherwise} \\
	\end{cases} 
\]

Then the residual graph corresponding to $f$ and $G$ is indicated by $G_f = (V, E_f)$, where the residual edge set $E_f$ is defined as following:
\[ E_f = \{(u,v) \in V \times V \colon c_f (u, v) > 0\}\]

After each flow update, the algorithm considers the residual graph corresponding to the new flow. Then the algorithm looks for an augmenting path--- a simple path from source to the sink--- in the residual graph. If such a path could be found, it will be used to construct a flow in the residual graph. Let $p$ be such a path. The residual capacity of $p$ is defined as $c_f(p) = min \{c_f(u, v): \text{$(u, v)$ is on  $p$}\}$. Then an augmenting flow $f_p \colon V \times V \rightarrow \mathbb{R}$  in the residual graph can be defined using $p$ as following:
\[ f_p (u, v) = 
	\begin{cases}
	c_f(p) \hfill & \text{ if $(u, v)$ is on $p$} \\	
	0 \hfill & \text{ otherwise} \\
	\end{cases} 
\]

This flow provides a road map for modifying the current flow in the flow network. The process of increasing the current flow using an augmenting flow is called augmentation. Let $f$ be the current flow in an $s$-$t$ network $G = (V, E)$ with capacity function $c$, and $f'$ be a flow in the corresponding residual graph. The augmentation of $f$ using $f'$ is a function from $V \times V$ to $\mathbb{R}$, defined as following:
\[ (f \uparrow f') (u, v) = 
	\begin{cases}
	f (u, v) + f'(u, v) - f'(v, u) \hfill & \text{ if $(u, v) \in E$} \\	
	0 \hfill & \text{ otherwise} \\
	\end{cases} 
\]

The intuition behind the above definition is based on the definition of the residual graph. The augment procedure increases the flow on $(u, v)$ by $f ( u, v )$ but decreases it by $f (v, u)$. This is due to the fact that pushing flow on the reverse edge on the residual graph signifies decreasing the flow in the original network.

In each iteration of the algorithm we replace the current flow with the result of the augment function--- another flow on the network but with greater value. The algorithm terminates if no more augmenting paths can be found in the residual graph. According to the Ford-Fulkerson theorem, the flow computed after termination is indeed the maximum flow in the flow network. According to the theorem, the following statements are equivalent:

\begin{itemize}
\item $f$ is a maximum $s$-$t$ flow in flow network $G$.
\item there is no augmenting path in the residual graph $G_f$.
\item there is an $s$-$t$ cut $C$ in $G$ such that capacity of $C$ is equal to the value of $f$.
\end{itemize}

The termination of the Ford-Fulkerson algorithm depends on how the augmenting path is found. For the networks with irrational edge capacities the algorithm might even fail to terminate~\cite{Zwick95}. In practice, the maximum-flow problem often arises with integral capacities. Moreover, we may convert rational edge capacities to integral values by multiplying them with a big enough integer. If $f^*$ denotes a maximum-flow in the transformed network, then the algorithm will iterate at most $|f^*|$ times, since the augmentation procedure increases the flow value by at least one unit in each iteration. Assuming that we use breadth-first-search (BFS) or depth-first-search (DFS) for finding augmenting paths makes the total execution time of the Ford-Fulkerson algorithm $O ( E | f^* |)$, as BFS and DFS run in $O (E + V)$.

(TODO: Here I mention the Edmonds-Karp algorithm and I mention its running time, co-relate this with definition in later sections) The Ford-Fulkerson algorithm is sometimes called a method instead of an algorithm because it does not fully specify the procedure for finding the augmenting path in the residual graph. There are several implementations with different execution times. For instance, the Edmond-Karp algorithm uses breadth-first-search (BFS) for finding augmenting paths and has $O (VE^2)$ running-time. The augmentings paths that are computed using BFS are also shortest paths connecting source to the sink in the residual graph. Edmond and Karp showed that in the Ford-Fulkerson algorithm, if each augmenting path is the shortest one, the length of the augmenting paths is non-decreasing and the algorithm always terminates. Using this statement they could finally present a polynomial algorithm for solving the maximum flow problem in generic case of real edge capacities.

\section{Formalizing the Ford-Fulkerson method}
%  GOAL: Present highlights of our formalization, persuade the reader that we have done something substantial.
%    Reader should think: Yeah, they did cool stuff!
%
%  [???* Thematize definition of flows: On Graphs vs. on Networks. You find both in the literature. 
%      We decides for [...]. This is better suited, as is gives nice and elegant proofs, even formally, bla bla bla]
%    
%  Proof of MinCut-MaxFlow: Follows the textbook proof.
%    Uses Isar to even look like a textbook proof, being comprehensible even without running
%      Isabelle/HOL.
%      
%    For example, \isai{augment_flow_presv_cap}.
%      Present formal proof text. Perhaps oppose it to textbook proof?.
%    
%  Abstract Algo looks like pseudocode presented in texbooks.
%    Oppose our algo to textbook pseudocode.
%

@@

(TODO: not sure about repeating citation of Cormen) Informal proofs focus on the relevant thoughts by leaving out technical details. However, in a formal proof each step has to be precisely specified as a valid application of some inference rule which makes formal proofs often lengthy and hard to follow. In the case that an interactive theorem prover is used, readers may even need to replay the proof to understand the idea behind it. In Isabelle proof assistant, the Isar proof language bridges the semantic gap between internal notions of proof given by interactive theorem proving systems and an appropriate level of abstraction for user-level work. Isar makes formal proofs readable even for those who are not familiar with Isabelle concepts. We have used such a functionality to present our formal proof of the Ford-Fulkerson method which resembels the informal proof described by Cormen et al.~\cite{CLRS09}.

As an example consider the following proof obtained form Cormen's algorithm book. Let $cf$ be the residual graph corresponding to a flow $f$ in an $s$-$t$ network $c$, and let $f'$ be a flow in $cf$. Consider the definition of the augment function that was described in previous section. Then, the result of augmenting $f$ with $f'$ is another function which satisfies the capacity constraint of a $s$-$t$ flow in $c$.

\emph{\textbf{Proof.}} first observe that if $(u, v) \in E$, then $c_f(v, u) = f(u, v)$. Therefore, we have $f'(v,u) \leq c_f(v, u) = f(u, v)$, and hence
	\begin{align*}
	(f \uparrow f') (u, v) &= f(u, v) + f'(u, v) - f'(v, u)  && \text{(by definition of augment)} \\
	& \geq f(u, v) + f'(u, v) - f(v, u) && \text{(because $f'(v,u) \leq f(u, v)$)} \\
	& = f'(u, v) \\
	& \geq 0.
	\end{align*}
In addition,
	\begin{align*}
	(f \uparrow f') (u, v) &= f(u, v) + f'(u, v) - f'(v, u)  && \text{(by definition of augment)} \\
	& \leq f(u, v) + f'(u, v) && \text{(because flows are nonnegative)} \\
	& \leq f(u, v) + c_f(u, v) &&  \text{(capacity constraint)} \\
	& = f(u, v) + c(u, v) - f(u, v) && \text{(definition of $c_f$)} \\
	& = c (u, v).
	\end{align*}

Following we present the equivalent Isar proof. Note that we have to additionally consider the case that $(u, v) \notin E$. Moreover in line 6, we have used the assumption that there are no parallel edges in a flow network. This important property is not mentioned in Cormen's proof.
\Snippet{augment_flow_presv_cap}

In reference books, it is a convention to use pseudocodes for presenting the algorithms. The abstract version of the algorithm has a clean correctness proof as it only captures the idea behind the algorithm and leaves out the implementation details. We present the Ford-Fulkerson method using the syntax of (TODO: here I mention autoref again, I was not aware that we are not using it anymore. So please correct all the references that I made to autoref tool) Autoref~\cite{La13} tool which is quite similar to pseudocode of the algorithm:
\Snippet{ford_fulkerson_algo}

In the pseudocode of the Ford-Fulkerson method, the path found while evaluating the loop condition is later used for augmenting the flow inside the loop body. The best way for implementing such requirement using syntax of the refinement framework is moving the path lookup inside the loop body. To do so, we modified the while-loop by introducing a break flag. The flag is initially off, and is set if no augmenting path is found in the residual graph.

\section{Refinement in Isabelle/HOL}
When formalizing algorithms, there is often a trade off between verifiability 
and efficiency: An efficient algorithm that uses elaborate data structures and optimizations is often harder to 
verify than a simple but straightforward algorithm.

The idea of stepwise refinement \cite{Wirth} is to first specify a simple, abstract version of an algorithm, which can be easily verified,
and then refine this algorithm towards an efficient implementation. Here, each refinement step can focus on a single aspect of the algorithm or 
its implementation, independently of the other refinement steps. This results in a greatly increased modularity of proofs. 
Refinement calculi \cite{Back} are used to systematically perform refinement in a Hoare-logic setting, and are particularly well-suited for 
implementation in theorem provers.

Note that it is important to support nondeterminism in a refinement setting, as this allows to defer 
implementation choices to the later refinement steps. For example, if we want to specify an algorithm 
that returns a shortest path between two nodes of a graph, we do not want to fix a particular shortest path.
Only later, when we decide to implement the specification by, e.g., breadth first search, we fix a particular path.

In Isabelle/HOL, stepwise refinement is supported by the Isabelle Refinement Framework~\cite{LaTu12}. 
It features a refinement calculus for programs phrased in a nondeterminism monad. 
The monad's type is a set of results plus an additional value that indicates a failure:
\begin{lstlisting}
  datatype \<alpha> nres = res \<alpha> set | fail
\end{lstlisting}
The return operation \isai{return x} of the monad describes the single result \isai{x}, and the bind 
operation \isai{bind m f} nondeterministically picks a result from $m$ and executes $f$ on it. 
If either \isai{m = fail}, or \isai{f} may fail for a result in $m$, the bind operation fails.

On nondeterministic results we define the \emph{refinement ordering} by lifting the subset ordering with \isai{fail} being the greatest element.
Intuitively, $m\le m'$ means that $m$ is a refinement of $m'$, i.e., all possible results of $m$ are 
also possible results of $m'$. 
Note that this ordering is a complete lattice, and bind is monotonic. Thus, we can define recursion using a fixed point construction~\cite{Kr10}.
Moreover, we can use the standard Isabelle/HOL constructs for if, let and case distinctions, yielding a fully fledged programming 
language, shallowly embedded into Isabelle/HOL's logic. For simpler usability, we define constants for loop constructs (while, foreach), 
assertions and specifications, and use a Haskell-like do-notation:
\begin{lstlisting}
  assert \<Phi> = if \<Phi> then return () else fail
  spec x. P x = RES {x. P x}
  do {x <- m; f x} = bind m f
  do {m; m'} = bind m (%_. m')
\end{lstlisting}

>>>>

\section{Introduction}
The maximum flow problem is an important problem in graph theory, as many other problems like the maximum-bipartite-matching problem, the edge-disjoint-paths problem, the circulation-demand problem, and various scheduling and resource allocating problems can be reduced to it.
The Ford-Fulkerson method~\cite{FF56} describes a class of algorithms to solve the maximum flow problem, an important instance being the Edmonds-Karp algorithm~\cite{EK72},
which solves the maximum flow problem in polynomial time for the general case of real valued capacities.

In this paper, we present a verified efficient implementation of the Edmonds-Karp algorithm and its polynomial complexity bound.
The formalization is conducted entirely in the Isabelle/HOL proof assistant~\cite{NPW02}. 
Stepwise refinement techniques~\cite{Wirth71,Back78,BaWr98} allow us to elegantly structure our verification into an abstract proof of the Ford-Fulkerson method,
its instantiation to Edmonds-Karp algorithm, and finally its efficient implementation. The abstract parts of our verification closely follow the standard textbook presentation of Cormen et al.~\cite{CLRS09}. Being developed in the Isar~\cite{Wenzel99} proof language, our proofs are accessible even to non-Isabelle experts.

The rest of this paper is structured as follows: In Section~\ref{sec:background} we give a short informal introduction to the Ford-Fulkerson method.
In Section~\ref{sec:abs-formalization}, we report on our formalization of the abstract method. 
Section~\ref{sec:refinement} gives a brief overview of the Isabelle Refinement Framework~\cite{LaTu12,La12}, which supports stepwise refinement based algorithm development in Isabelle/HOL. In Section~\ref{sec:edka}, we report on our instantiation of the Ford-Fulkerson method to the Edmonds-Karp algorithm and the proof of its complexity.
Section~\ref{sec:executable} reports on the further refinement steps required to yield an efficient implementation. Section~\ref{sec:benchmark} reports on 
benchmarking our implementation against a reference implementation of the Edmonds-Karp algorithm from Sedgewick et al.~\cite{SeWa11}. 
Finally, Section~\ref{sec:concl} gives a conclusion and discusses related and future work.

% Despite its importance, no formalization of the Ford-Fulkerson algorithm has been developed in modern proof assistants. The only similar work that the authors are aware of is formalization of the Ford-Fulkerson algorithm in Mizar~\cite{Lee05}. This formalization defines and proves correctness of the algorithm at the level of graph manipulations without providing concrete implementation of the algorithm. Providing such an implementation is specially important, as it could provide the basis for many verfied programs with practical importance.

\section{A short background on the Ford-Fulkerson method}\label{sec:background}
% \begin{comment}
%   GOAL: 
%     1) Remind the reader of FoFu. The educated reader should have the feeling:
%       Yes, now I recall FoFu, and understand (again) what it does.
%       
%     2) Set the field for further, more detailed descriptions in next section.
%   
% 
%   Short description of Network: Finite graph over nodes V, 
%     edge (u,v) annotated with capacity c(u,v).
%     Assuming distinct nodes s and t. 
%     Note on additional assumptions: 
%         No parallel edges, s only outgoing, t only incoming, all nodes on path from s to t. 
%         Can transform any Network to match these assumptions, preserving the flow \cite{???}.
%         
%   s-t Flow: Annotation of edges with values, such that: 
%     1) Values smaller than capacities.
%     2) Kirchhoff law: sum of incoming flows = sum of outgoing flows, for all nodes except s t
%   
%   Value of flow: Incoming flow - outgoing flow of s
%   
%   Max-Flow problem. 
%     Rpt. why important?
%   
%   Cuts, intuition of min-cut >= max-flow
%   
%   Theorem: Min-Cut = Max-Flow
%   Proven via augmenting flow in residual graph. 
%     Present 1,2,3 of min-cut max-flow equivalences
%   
%   This yields an algorithm to compute max-flows, if we can compute flows in residual graph.
%     --> Simple way: Augmenting flow via augmenting path. ==> FoFu-scheme. 
%       Termination? In General: Only for integer capacities.
%       Edmonds-Karp: Shortest augmenting path. Always terminates! Complexity: O(|E||V|) outer loop iterations, BFS requires O|E|.
%   \end{comment}

In this section, we give a short introduction to the Ford-Fulkerson method, closely following the presentation in~\cite{CLRS09}.

A (flow) network is a directed graph over a finite set of vertices $V$ and edges $E$, where each edge $(u,v)\in E$ is labeled by a positive real valued capacity $c(u,v)>0$.
Moreover, their are two distinct vertices $s,t\in E$ which are called \emph{source} and \emph{sink}. To simplify reasoning about flow networks, it is common to make some additional assumptions: 1) The source only has outgoing edges while the sink only has incoming edges; 2) If the flow network contains an edge $(u, v)$ then there is no edge $(v, u)$ in the reverse direction; and 3) every vertex of the flow network must be on a path from $s$ to $t$. Although these assumptions seem restrictive, any network may be transformed such that it satisfies all the aforementioned properties~\cite{CLRS09}.

An flow $f$ on a network is a labeling of the edges with real values satisfying the following constraints: 1) \emph{Capacity constraint}: the flow on each edge is a non-negative value smaller or equal to the edge's capacity; 2) \emph{Conservation constraint}: For all vertices except $s$ and $t$, the sum of flows over all incoming edges is equal to the sum of flows over all outgoing edges.

The value of flow $f$ is denoted by $|f|$, and defined to be the sum over the flows of all outgoing edges of the source\footnote{Note that we assumed that the source has no incoming edges. In a more general setting, the value of a flow is the sum over all flows leaving $s$ minus the sum over all incoming flows to $s$.}.
Given a network $G$, the maximum flow problem is to find a flow with a maximum value among all flows of the network.

An important result is the relation between flows and cuts in a network. A \emph{cut} in a network is a partition of the vertices into two sets, such that one set contains 
the source and the other set contains the sink. The capacity of a cut is the sum of the capacities of all edges going from the source's side to the sink's side of the cut.
It is easy to see that the value of any flow cannot exceed the capacity of any cut, as all flow from the source must ultimately reach the sink, and thus go through the edges of the cut. The Ford-Fulkerson theorem tightens this bound and states that the value of the maximum flow is equal to the capacity of the minimum cut.

The Ford-Fulkerson method is a corollary of this theorem. It is based on a greedy approach: Starting from a zero flow, the value of the flow is iteratively increased until a maximum flow is reached. In order to increase the flow value, it may be necessary to redirect some flow, \ie to decrease the flow passed through specific edges. For this purpose, the Ford-Fulkerson method defines the residual graph. It has edges in parallel and reverse to the edges of the network, each edge indicating how much flow can be moved between two nodes by either increasing or decreasing the flow on an edge in the network. Formally, the residual graph $G_f$ of a flow $f$ is the graph induced by the 
edges with positive labels according to the following labeling function $c_f$:
\[ c_f (u, v) = 
  \begin{cases}
  c (u, v) - f(u, v) \hfill & \text{ if $(u, v) \in E$} \\
  f (v, u) \hfill & \text{ if $(v, u) \in E$} \\  
  0 \hfill & \text{ otherwise} \\
  \end{cases} 
\]

In each iteration, the Ford-Fulkerson method tries to find an \emph{augmenting path}, \ie a simple path from $s$ to $t$ in the residual graph, and then 
pushes as much flow as possible along this path to increase the value of the current flow. 
Formally, for an augmenting path $p$, one defines the \emph{residual capacity} $c_p$ as the minimum value over all edges of $p$:
\[c_f(p) = \min \{c_f(u, v): \text{$(u, v)$ is on  $p$}\}\]
An augmenting path yields a residual flow $f_p$, which is the flow that can be passed along this path:
\[ f_p (u, v) = 
  \begin{cases}
  c_f(p) \hfill & \text{ if $(u, v)$ is on $p$} \\  
  0 \hfill & \text{ otherwise} \\
  \end{cases} 
\]
\newcommand{\augment}{{\mathbin\uparrow}}
In order to actually push the flow induced by an augmenting path, we define the \emph{augment} function $f\augment f'$, which augments a flow $f$ in the network 
by any \emph{augmenting flow} $f'$, i.e., any flow in the residual graph:
\[ (f \augment f') (u, v) = 
  \begin{cases}
  f (u, v) + f'(u, v) - f'(v, u) \hfill & \text{ if $(u, v) \in E$} \\  
  0 \hfill & \text{ otherwise} \\
  \end{cases} 
\]
Note that, for any edge in the network, the augmenting flow parallel to that edge is added to the flow, while the augmenting flow reverse to this edge is subtracted. 
This matches the intuition of passing flow in the indicated direction, by either increasing or decreasing the flow of an edge in the network.

The correctness of the Ford-Fulkerson algorithm follows from the Ford-Fulkerson theorem, which is usually stated as the following three statements being equivalent:
\begin{enumerate}
\item $f$ is a maximum flow in a network $G$.
\item there is no augmenting path in the residual graph $G_f$.
\item there is a cut $C$ in $G$ such that the capacity of $C$ is equal to the value of $f$.
\end{enumerate}

The Ford-Fulkerson method does not specify how to find an augmenting path in the residual graph, and there are several possible implementations with different execution times. The general method is only guaranteed to terminate for networks with rational capacities, while it may infinitely converge against non-maximal flows in the case of irrational edge capacities~\cite{FF56,Zwick95}. When always choosing a \emph{shortest} augmenting path, the number of iterations is bound by $O(VE)$\footnote{Where unambiguous, we write $V$ and $E$ instead of $|V|$ and $|E|$ for the number of nodes and edges.}, even for the general case of real valued capacities. A shortest path can be found by breadth first search (BFS) in time $O(E)$, yielding the Edmonds-Karp algorithm~\cite{EK72} with an overall running time of $O(VE^2)$.

\section{Formalizing the Ford-Fulkerson method}\label{sec:abs-formalization}
%  GOAL: Present highlights of our formalization, persuade the reader that we have done something substantial.
%    Reader should think: Yeah, they did cool stuff!
%
%  [???* Thematize definition of flows: On Graphs vs. on Networks. You find both in the literature. 
%      We decides for [...]. This is better suited, as is gives nice and elegant proofs, even formally, bla bla bla]
%    
%  Proof of MinCut-MaxFlow: Follows the textbook proof.
%    Uses Isar to even look like a textbook proof, being comprehensible even without running
%      Isabelle/HOL.
%      
%    For example, \isai{augment_flow_presv_cap}.
%      Present formal proof text. Perhaps oppose it to textbook proof?.
%    
%  Abstract Algo looks like pseudocode presented in texbooks.
%    Oppose our algo to textbook pseudocode.
%

@@

% (DONE: not sure about repeating citation of Cormen: I would definitely repeat the citation!) 
Informal proofs focus on the relevant thoughts by leaving out technical details and obvious steps. 
In contrast, a formal proof has to precisely specify each step as the application of some inference rules. 
Although modern proof assistants provide high-level tactics to summarize some of these steps, formal proofs tend to be significantly more 
verbose than informal proofs. Moreover, formal proofs are conducted in the tactic language of the proof assistant, which is often some dialect of ML. 
Thus, many formal proofs are essentially programs that instruct the proof assistant how to conduct the proof, and are inaccessible without a deep knowledge of the used proof assistant, in many cases requiring to replay the proof in the proof assistant in order to understand the idea behind it.

For the Isabelle/HOL proof assistant, the Isar proof language~\cite{Wenzel99} allows to write formal proofs that strongly resemble standard mathematical textbook proofs, and are accessible even for those not familiar with Isabelle/HOL. 
We use Isar to present our proof of the Ford-Fulkerson method such that it resembles the informal proof described by Cormen et al.~\cite{CLRS09}.

As an example consider the proof that augmenting a flow $f$ with a residual flow $f'$ yields a valid flow $f\augment f'$ again. In particular, one has to show 
that the augmented flow satisfies the capacity constraint. Cormen et al.~give the following proof, which we copied literally, only replacing the references to 
``Equation~26.4'' by ``definition of $\augment$'':

For the capacity constraint, first observe that if $(u, v) \in E$, then $c_f(v, u) = f(u, v)$. Therefore, we have $f'(v,u) \leq c_f(v, u) = f(u, v)$, and hence
	\begin{align*}
	(f \uparrow f') (u, v) &= f(u, v) + f'(u, v) - f'(v, u)  && \text{(by definition of $\augment$)} \\
	& \geq f(u, v) + f'(u, v) - f(v, u) && \text{(because $f'(v,u) \leq f(u, v)$)} \\
	& = f'(u, v) \\
	& \geq 0.
	\end{align*}
In addition,
	\begin{align*}
	(f \uparrow f') (u, v) &= f(u, v) + f'(u, v) - f'(v, u)  && \text{(by definition of $\augment$)} \\
	& \leq f(u, v) + f'(u, v) && \text{(because flows are nonnegative)} \\
	& \leq f(u, v) + c_f(u, v) &&  \text{(capacity constraint)} \\
	& = f(u, v) + c(u, v) - f(u, v) && \text{(definition of $c_f$)} \\
	& = c (u, v).
	\end{align*}

In the following we present the corresponding formal proof in Isar:
\Snippet{augment_flow_presv_cap}
Its structure is exactly the same as the textbook proof, except that we had to also consider the case $(u,v)\notin E$, which is not mentioned in the informal proof at all, and easily discharged in our formal proof by the \isai{auto}-tactic after the \isai{qed}. We also use exactly the same justifications as the original proof, except that we had to use the fact that there are no parallel edges to show $c_f(v,u)=f(u,v)$, which is not mentioned in the original proof.

In textbooks, it is common to present algorithms in pseudocode, which captures the essential ideas, but leaves open implementation details. As a formal equivalent to pseudocode, we use the monadic programming language provided by the Isabelle Refinement Framework~\cite{LaTu12,La12}:
\Snippet{ford_fulkerson_algo}

TODO: Introduce obtain-syntax, and unfold find-augmenting-path spec!
The code looks quite similar to pseudocode that one would expect in a textbook, but actually is a rigorous formal specification of the algorithm, using nondeterminism to leave open the implementation details (\cf Section~\ref{sec:refinement}). The only caveat is that our framework currently does not support breaking of loops, such that we had to use an explicit flag instead.

Finally, using the Ford-Fulkerson theorem and the verification condition generator of the Refinement Framework, it is straightforward to prove (partial) correctness of
the Ford-Fulkerson method, which is stated in Isabelle/HOL by the following theorem:
\Snippet{ford_fulkerson_partial_correct}

\section{Refinement in Isabelle/HOL}\label{sec:refinement}
After having stated and proved correct an algorithm on the abstract level, the next step is to provide an efficient implementation. 
In our case, we first specialize the Ford-Fulkerson method to use shortest augmenting paths, then implement the search for shortest augmenting paths by BFS, and finally use efficient data structures to represent the abstract objects modified by the algorithm.

A natural way to achieve this formally is \emph{stepwise refinement}~\cite{Wirth71}, and in particular \emph{refinement calculus}~\cite{Back78,BaWr98}, which allows us to systematically transform an abstract algorithm into a more concrete one, preserving its correctness.

In Isabelle/HOL, stepwise refinement is supported by the Isabelle Refinement Framework~\cite{LaTu12}. 
It features a refinement calculus for programs phrased in a nondeterminism monad. 
The monad's type is a set of possible results plus an additional value that indicates a failure:
\begin{lstlisting}
  datatype \<alpha> nres = res \<alpha> set | fail
\end{lstlisting}
The operation \isai{return x} of the monad describes the single result \isai{x}, and the 
operation \isai{bind m f} nondeterministically picks a result from $m$ and executes $f$ on it. 
If either \isai{m = fail}, or \isai{f} may fail for a result in $m$, the bind operation fails.

We define the \emph{refinement ordering} on \isai{\<alpha> nres} by lifting the subset ordering with \isai{fail} being the greatest element.
Intuitively, $m\le m'$ means that $m$ is a refinement of $m'$, i.e., all possible results of $m$ are 
also possible results of $m'$. 
Note that the refinement ordering is a complete lattice, and bind is monotonic. Thus, we can define recursion using a fixed point construction~\cite{Kr10}.
Moreover, we can use the standard Isabelle/HOL constructs for if, let and case distinctions, yielding a fully fledged programming 
language, shallowly embedded into Isabelle/HOL's logic. For simpler usability, we define constants for loop constructs (while, foreach), 
assertions and specifications, and use a Haskell-like do-notation:
\begin{lstlisting}
  assert \<Phi> = if \<Phi> then return () else fail
  spec x. P x = RES {x. P x}
  do {x <- m; f x} = bind m f
  do {m; m'} = bind m (%_. m')
\end{lstlisting}
